Let's fully get set up first by importing and looking at the data structure:
```{r}
rm(list = ls())
library(lme4)
library(tidyverse)
library(zoo)
library(ggeffects)
library(performance)
library(sjPlot)
library(influence.ME)
library(ggsci)

phenology <- read.csv("../data/phenology(1).csv")
str(phenology)
head(phenology)

girth <- read.csv("../data/girth(1).csv")
str(girth)
head(girth)

#Metadata for weather data
weather_station <- read.delim("../data/WeatherStation.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

trees <- read.csv("../data/trees(1).csv")
str(trees)
head(trees)

SilwoodWeatherDaily <- read.csv("../data/SilwoodWeatherDaily(1).csv")
str(SilwoodWeatherDaily)
head(SilwoodWeatherDaily)
```
The .csv files that can easily be combined into one df are phenology, trees and girth, so I'll just do that for now:
```{r}
# merge phenology and trees first by the common column "TreeID"
budburst <- merge(phenology, trees, by = "TreeID")
str(budburst)
head(budburst)
```
Potential RQ: How does air temperature and rainfall (mm) affect date of budding in Quercus rober in Silwood Park?
Hypothesis: Increases in air temperature and rainfall (mm) cause Quercus rober trees to bud earlier in the year
NOTE: Data for 1 April to 25th September has been recorded using a different rain gauge
Now we need to merge the temperature and rainfall with budburst using the correct dates (this will create replicates that we remove later by looking for the earliest day of budburst)
```{r}
SilwoodWeatherDaily <- SilwoodWeatherDaily %>% mutate(year = format(as.Date(TIMESTAMP, format = "%d/%m/%Y"), "%Y"), julian_day = format(as.Date(TIMESTAMP, format = "%d/%m/%y"), "%j"))
str(SilwoodWeatherDaily)

budburst <- budburst %>% mutate(year = format(as.Date(date, format = "%d/%m/%Y"), "%Y"), julian_day = format(as.Date(date, format = "%d/%m/%y"), "%j"))
str(budburst)

# merge by julian day and year only
budburst_final <- merge(budburst, SilwoodWeatherDaily, by = c("julian_day", "year"))
str(budburst_final)
head(budburst_final)
```
Now that the merger is successful, we have a dataset of Tree IDs by year and julian day, where there are scores of budburst for day and each tree
The time has come for some data-wrangling, starting by removing the not required species, then only keeping the valid scores of budburst, then using slice_min to find the earliest budburst julian day for each TreeID per year
```{r}
budburst_final <- budburst_final %>% filter(species == "quercus.robur") %>% filter((score %in% c("1", ">1"))) 
budburst_final <- budburst_final %>% group_by(TreeID, year) %>% slice_min(order_by = julian_day, n = 1) %>% ungroup()
str(budburst_final)
head(budburst_final)
```
Ok, so now the data is finally filtered correctly to include the first instance of budburst/greening in each oak tree per year
But there is a problem, we would only be left with data for the temperature and rainfall for the day of budding, and no other day before, which is useless
To counter this problem, we must calculate 7 day and 14 day rolling means for rainfall using the zoo package, and also calculate cum GDD as a proxy for rolling mean of temperature relative to the base temperature of oak
```{r}
# Calculate GDD with base temperature 5°C (common for oak)

base_temp <- 5

SilwoodWeatherDaily <- SilwoodWeatherDaily %>%
    mutate(
        mean_temp = (Air_Temp_Max...Deg.C...Max. + Air_Temp_Min...Deg.C...Min.) / 2,
        GDD = pmax(0, mean_temp - base_temp)) %>%
    group_by(year) %>%
    mutate(cumGDD = cumsum(GDD)) %>%
    ungroup()

SilwoodWeatherDaily <- SilwoodWeatherDaily %>%
    arrange(julian_day) %>%
    group_by(year) %>%
    mutate(
    #If cum GDD is deemed useless, then can also calculate 7 day or 14 day means for temperature
    mean_temp_7d = rollmean(mean_temp, k = 7, fill = NA, align = "right"),
    mean_temp_14d = rollmean(mean_temp, k = 14, fill = NA, align = "right"),
    rain_7d = rollsum(Rain_mm_Tot...mm...Tot., k = 7, fill = NA, align = "right"),
    rain_14d = rollsum(Rain_mm_Tot...mm...Tot., k = 14, fill = NA, align = "right")
) %>%
    ungroup()

budburst_env <- merge(
    budburst_final,
    SilwoodWeatherDaily[, c("julian_day", "year", "mean_temp", "mean_temp_7d", "mean_temp_14d",
    "rainfall" = "Rain_mm_Tot...mm...Tot.", "rain_7d", "rain_14d", "cumGDD")],
    by = c("julian_day", "year"),
    all.x = TRUE
)
str(budburst_env)
head(budburst_env)
budburst_env$TreeID <- as.factor(budburst_env$year)
budburst_env$julian_day <- as.numeric(budburst_env$julian_day)
```
Now what is left to do is to perform a linear mixed-model with temperature, rainfall, and their interaction as the predictors, and TreeID, Year and Location as the random effects
I'm also going to try some variations to see if my model can be improved in any way:
```{r}
#trial1: scaled 14 day rainfall with ID and year as random effects
trial1 <- lmer(formula = julian_day ~ scale(rain_14d) + (1 | TreeID) + (1 | year), data = budburst_env)
summary(trial1)
#trial1 without random effects just for model validation
trial1 <- lm(formula = julian_day ~ scale(rain_14d), data = budburst_env)
summary(trial1)

#trial2: scaled GDD and scaled 14 day rainfall with ID, year and location as random effects
trial2 <- lmer(formula = julian_day ~ scale(cumGDD) * scale(rain_14d) + (1 | TreeID) + (1 | year) + (1 | SPlocation), data = budburst_env)
summary(trial2)

#trial3: scaled GDD and 7 day rainfall with ID, year and location as random effects 
trial3 <- lmer(formula = julian_day ~ scale(cumGDD) * scale(rain_7d) + (1 | TreeID) + (1 | year) + (1 | SPlocation), data = budburst_env)
summary(trial3)

par(mfrow = c(2,2))
plot(M1)
```
Graphing: we are sticking to M1 for now, which models julian day of first budburst as the response variable explained by z-scaled 14 day prior mean rainfall values
For this, we create one plot with one slope for all random effects averages, and color-coding by the random effect which is year
```{r}
# Fit model with the scale
budburst_env <- budburst_env %>%
  mutate(rain_14d_scaled = scale(rain_14d)[,1])

M1 <- lmer(julian_day ~ rain_14d_scaled + (1 | TreeID) + (1 | year) + (1 | SPlocation), data = budburst_env)
summary(M1)

pred_M1 <- ggpredict(M1, terms = "rain_14d_scaled [all]", type = "fixed")

# Plot
ggplot(budburst_env, aes(x = rain_14d_scaled, y = julian_day, color = factor(year))) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.6, size = 2) +
  geom_line(data = pred_M1, aes(x = x, y = predicted), color = "black", linewidth = 1.2) +
  geom_ribbon(data = pred_M1, 
              aes(x = x, ymin = conf.low, ymax = conf.high), 
              fill = "grey70", alpha = 0.3, inherit.aes = FALSE) +
  labs(
    x = "Scaled sum of rainfall 14 days prior to budburst",
    y = "Julian day of first budburst",
    color = "Year"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "right")
```
Now this graph is fine, it shows summed rainfall 14 days prior to budburst day, and is color-coded by year
Only one problem: the values are still zero-inflated, meaning that they are close to 0 when not scaled, even with rollsum
There are two ways to counter this, the first of which is logging and adding 1 using log1p, then scaling it:
```{r}
#Method 1: logging and adding 1
budburst_env <- budburst_env %>%
  mutate(rain_14d_log = log1p(rain_14d))
str(budburst_env)

budburst_env <- budburst_env %>%
  mutate(rain_14d_log_scaled = scale(rain_14d_log)[,1])

M2 <- lmer(julian_day ~ rain_14d_log + (1 | TreeID) + (1 | year), data = budburst_env)
summary(M2)

pred_M2 <- ggpredict(M2, terms = "rain_14d_log [all]", type = "fixed")

# Plot
cb11_noredgreen <- c(
  "#332288",  # dark blue-purple
  "#88CCEE",  # light cyan-blue
  "#44AAEE",  # mid blue
  "#1178B4",  # strong blue
  "#DDCC77",  # muted yellow
  "#AA4499",  # magenta-purple
  "#CC99BB",  # light magenta
  "#8888AA",  # gray-blue
  "#A1887F",  # brown-gray
  "#6A4C93",  # violet
  "#B07AA1"   # dusty lavender
)

ggplot(budburst_env, aes(x = rain_14d_log, y = julian_day, color = factor(year))) +
  geom_jitter(width = 0.08, height = 0, alpha = 0.6, size = 2) +
  geom_line(data = pred_M2, aes(x = x, y = predicted), color = "black", linewidth = 1.2) +
  geom_ribbon(
    data = pred_M2,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    fill = "grey70", alpha = 0.3, inherit.aes = FALSE
  ) +
  labs(
    x = "Log(sum of rainfall 14 days prior to budburst + 1)",
    y = "Julian day of first budburst",
    color = "Year"
  ) +
  scale_color_manual(values = cb11_noredgreen) +
  theme_bw(base_size = 13) +
  theme(legend.position = "right")


##NOW JUST FOR DIAGNOSTIC PLOTS##

#M1 <- lm(julian_day ~ rain_14d_log_scaled, data = budburst_env)

#par(mfrow = c(2,2))
#plot(M1)
```
The second way to work with zero-inflated values is to categorize them into high and low rainfall from the median:
```{r}
#Method 2: Categorizing 14 day rainfall sum into high and low
median_rain <- median(budburst_env$rain_14d, na.rm = TRUE)

budburst_env <- budburst_env %>%
  mutate(rain_cat = ifelse(rain_14d > median_rain, "High", "Low"))

M3 <- lmer(julian_day ~ rain_cat + (1 | TreeID) + (1 | year), data = budburst_env)
summary(M3)

#Addition: 3 categories instead of 2
budburst_env <- budburst_env %>%
  mutate(rain_cat3 = cut(
    rain_14d,
    breaks = quantile(rain_14d, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
    include.lowest = TRUE,
    labels = c("Low", "Medium", "High")
  ))

M4 <- lmer(julian_day ~ rain_cat3 + (1 | TreeID) + (1 | year), data = budburst_env)
summary(M4)

#Plot
ggplot(budburst_env, aes(x = rain_cat3, y = julian_day)) +
  geom_boxplot(fill = "grey85", color = "grey50", trim = FALSE, width = 0.8) +
  geom_jitter(aes(color = factor(year)),
              width = 0.15, alpha = 0.7, size = 2.2) +
  labs(
    x = "14-day rainfall category",
    y = "Julian day of first budburst",
    color = "Year"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank(),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10))
  )
```
I now have some trial models 1(same as M1) to 3, and I have my rainfall models, the continuos M1 and M2, and the categorical M3 and M4
Now to find out which model fits the data best and then stick to it forever:
```{r}
#Need to first add all the columns used into a new dataframe
common_data <- budburst_env %>%
  drop_na(rain_14d, rain_14d_scaled, rain_cat3, rain_cat, rain_14d_log, rain_14d_log_scaled, julian_day, cumGDD, rain_7d)
str(common_data)

trial2 <- lmer(formula = julian_day ~ scale(cumGDD) * scale(rain_14d) + (1 | TreeID) + (1 | year) + (1 | SPlocation), data = common_data)
trial3 <- lmer(formula = julian_day ~ scale(cumGDD) * scale(rain_7d) + (1 | TreeID) + (1 | year) + (1 | SPlocation), data = common_data)
M1 <- lmer(julian_day ~ rain_14d_scaled + (1 | TreeID) + (1 | year), data = common_data)
M2 <- lmer(julian_day ~ rain_14d_log + (1 | TreeID) + (1 | year), data = common_data)
M3 <- lmer(julian_day ~ rain_cat + (1 | TreeID) + (1 | year), data = common_data)
M4 <- lmer(julian_day ~ rain_cat3 + (1 | TreeID) + (1 | year), data = common_data)

anova(M1, M2)
compare_performance(M1, M2, M3, M4, rank = TRUE)
```
In terms of trials, trial3 is better than trial 2, and M4 is better than all other M rainfall models, but followed by M1
Now for M1, M2 and M4, I want the tab_model to take a look at the coefficients:
```{r}
tab_model(
  M1, M2, M4,
  show.re.var = TRUE,       # show random effect variances
  show.icc = TRUE,          # show intraclass correlation
  show.obs = TRUE,          # number of observations
  show.r2 = TRUE,           # marginal & conditional R²
  dv.labels = c("Model 1: Rain (scaled continuous)",
                "Model 2: Rain (log-scaled continuous)",
                "Model 4: Rain (categorical 3 levels)"),
  title = "Comparison of mixed-effects models for budburst timing"
)
```
Calculating cook's distance for an lmer:
```{r}
# Example: compute influence measures per TreeID
infl_M2 <- influence(M2, obs = TRUE)   # obs = TRUE gives case-level influence

# Calculate Cook's distance
cooks_M2 <- cooks.distance(infl_M2)

# Convert to data frame for plotting
cooks_df <- data.frame(
  obs = 1:length(cooks_M2),
  cooks = cooks_M2
)

# Identify those with Cook's D > 1
cooks_df$outlier <- cooks_df$cooks > 1

# Plot Cook’s distance
ggplot(cooks_df, aes(x = obs, y = cooks)) +
  geom_bar(stat = "identity", aes(fill = outlier)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  labs(
    x = "Observation",
    y = "Cook's distance",
    title = "Cook’s Distance for LMM (M2)"
  ) +
  scale_fill_manual(values = c("FALSE" = "grey70", "TRUE" = "tomato")) +
  theme_minimal(base_size = 13)
```